---
title: "03AttributeDataOperations"
output: html_document
date: "2025-03-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Attribute Data Operations

```{r}
library(sf)
library(dplyr)
library(spData)
library(terra)
```

# Attribute Data Manipulation

## Vector attributes

```{r}
methods(class = 'sf')
```

```{r}
class(world)
dim(world)
```

> Dropping the geometry

```{r}
world_df = st_drop_geometry(world)
class(world_df)
ncol(world_df)
```

## Vector Attribute Aggregation

> Only returns DataFrame

```{r}
world_agg1 = aggregate(pop ~ continent, FUN=sum, data=world,
                       na.rm = TRUE)
class(world_agg1)
```

> Returns geometry as well

```{r}
world_agg2 = aggregate(world["pop"], by = list(world$continent), FUN=sum,
                       na.rm=T)
class(world_agg2)
nrow(world_agg2)
```

```{r}
world_agg3 = world |> 
  group_by(continent) |> 
  summarize(pop = sum(pop, na.rm = T))
class(world_agg3)
```

```{r}
world_agg4 = world |> 
  group_by(continent) |> 
  summarize(Pop = sum(pop, na.rm = T),
            Area = sum(area_km2), 
            N = n())
class(world_agg4)
names(world_agg4)
```

```{r}
world_agg5 = world |> 
  st_drop_geometry() |> 
  select(pop, continent, area_km2) |> 
  group_by(Continent = continent) |> 
  summarise(Pop = sum(pop, na.rm=T), 
            Area = sum(area_km2), 
            N = n()) |> 
  mutate(Density = round(Pop / Area)) |> 
  slice_max(Pop, n = 3) |> 
  arrange(desc(N))
world_agg5
```

## Vector Attribute Joining

```{r}
world_coffee = left_join(world, coffee_data)
class(world_coffee)
```

```{r}
names(world_coffee)
plot(world_coffee["coffee_production_2017"])
```

```{r}
coffee_renamed = rename(coffee_data, nm = 'name_long')
world_coffee2 = left_join(world, coffee_renamed, 
                          by = join_by(name_long == nm))
```

> Can use `inner_join` to only keep matching rows.
>
> Find the difference between list of values in frame

```{r}
setdiff(coffee_data$name_long, world$name_long)
```

```{r}
drc = stringr::str_subset(world$name_long, "Dem*.+Congo")
drc
```

```{r}
coffee_data$name_long[grepl("Congo,", coffee_data$name_long)] = drc
world_coffee_match = inner_join(world, coffee_data)
nrow(world_coffee_match)
```

## Adding attributes

> Using base R

```{r}
world_new = world
world_new$pop_dens = world_new$pop / world_new$area_km2
```

> Using `dplyr`

```{r}
world_new2 = world |> 
  mutate(pop_dens = pop / area_km2)
```

> Using `transmute()` instead drops all other columns (except geom)

> Combining columns

```{r}
world_united = world |>
  tidyr::unite("con_reg", continent:region_un, sep = ":", remove = TRUE)
names(world_united)
```

```{r}
world_united[10,c('con_reg')]
```

> Splitting them

```{r}
world_separate = world_united |>
  tidyr::separate(con_reg, c("continent", "region_un"), sep = ":")
```

# Raster Objects

```{r}
elev = rast(nrows = 6, ncols = 6,
            xmin = -1.5, xmax = 1.5, ymin = -1.5, ymax = 1.5,
            vals = 1:36)
elev
```

```{r}
plot(elev)
```

> Add categorical data

```{r}
grain_order = c("clay", "silt", "sand")
grain_char = sample(grain_order, 36, replace = TRUE)
grain_fact = factor(grain_char, levels = grain_order)
grain = rast(nrows = 6, ncols = 6, 
             xmin = -1.5, xmax = 1.5, ymin = -1.5, ymax = 1.5,
             vals = grain_fact)
cats(grain)
```

```{r}
levels(grain)
```

```{r}
grain2 = grain # do not overwrite the original data
levels(grain2) = data.frame(value = c(0, 1, 2), wetness = c("wet", "moist", "dry"))
levels(grain2)
```

> Summarizing

```{r}
summary(grain)
```

> Get a table

```{r}
freq(grain)
```
