---
title: "04SpatialDataOperations"
output: html_document
date: "2025-03-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Spatial Data Operations - Vector Data

```{r}
library(sf)
library(terra)
library(spData)
library(dplyr)
library(tmap)
```

## Subsetting

```{r}
canterbury = nz |> filter(Name == "Canterbury")
canterbury_height = nz_height[canterbury, ]
```

```{r}
p_hpnz1 = tm_shape(nz) + 
  tm_polygons(fill = "white") +
  tm_shape(nz_height) + 
  tm_symbols(shape = 2, col = "red", size = 0.5, col_alpha = 0.75) +
  tm_title("High points in New Zealand") +
  tm_layout(bg.color = "lightblue")
p_hpnz2 = tm_shape(nz) + 
  tm_polygons(fill = "white") +
  tm_shape(canterbury) +
  tm_fill(col = "gray") + 
  tm_shape(canterbury_height) + 
  tm_symbols(shape = 2, col = "red", size = 0.5, col_alpha = 0.75) +
  tm_title("High points in Canterbury") +
  tm_layout(bg.color = "lightblue")
tmap_arrange(p_hpnz1, p_hpnz2, ncol = 2)

```

```{r}
canterbury_height[1,]
```

> Points that do not intersect

```{r}
nz_height[canterbury, , op = st_disjoint]
```

> Alternatively
>
> Use a *sparse geometry binary predicate*, convert it to a logical vector with T/F values

```{r}
sel_sgbp = st_intersects(x = nz_height, y = canterbury)
class(sel_sgbp)
sel_sgbp
```

```{r}
sel_logical = lengths(sel_sgbp) > 0
canterbury_height2 = nz_height[sel_logical, ]
canterbury_height2
```

> Alternatively again

```{r}
canterbury_height3 = nz_height |> 
  st_filter(y = canterbury, .predicate = st_intersects)
canterbury_height3
```

## Topological Relationships

> Binary

```{r}
polygon_matrix = cbind( 
  x = c(0, 0, 1, 1, 0),
  y = c(0, 1, 1, 0.5, 0)
)
polygon_sfc = st_sfc(st_polygon(list(polygon_matrix)))
polygon_sfc
```

> From a data frame

```{r}
point_df = data.frame(
  x = c(0.2, 0.7, 0.4),
  y = c(0.1, 0.2, 0.8)
)
point_sf = st_as_sf(point_df, coords = c("x", "y"))
point_sf
```

```{r}
par(pty = "s")
plot(polygon_sfc, border = "red", col = "gray", axes = TRUE, reset = FALSE)
plot(point_sf, add = TRUE, lab = 1:4, cex = 2)
text(point_df[, 1] + 0.02, point_df[, 2] + 0.04, 1:3, cex = 1.3)
```

> Which points intersect in some way with the polygon

```{r}
st_intersects(point_sf, polygon_sfc)
```

```{r}
st_intersects(point_sf, polygon_sfc, sparse = F)
```

```{r}
st_within(point_sf, polygon_sfc)
st_touches(point_sf, polygon_sfc)
```

```{r}
st_disjoint(point_sf, polygon_sfc, sparse = F)[,1]
```

```{r}
st_is_within_distance(point_sf, polygon_sfc, 
                      dist = 0.2, sparse = F)[,1]
```

## Distance relations

> Continuous

```{r}
nz_highest = nz_height |> slice_max(n = 1, order_by = elevation)
canterbury_centroid = st_centroid(canterbury)
st_distance(nz_highest, canterbury_centroid)
```

```{r}
co = filter(nz, grepl("Canter|Otag", Name))
st_distance(nz_height[1:3, ], co)
```

```{r}
plot(st_geometry(co)[2])
plot(st_geometry(nz_height)[2:3], add = T)
```

## Spatial joining

> Create random points on the Earth

```{r}
set.seed(2018)
(bb = st_bbox(world))
random_df = data.frame(
  x = runif(n = 10, min = bb[1], max = bb[3]), 
  y = runif(n = 10, min = bb[2], max = bb[4])
)
random_points = random_df |> 
  st_as_sf(coords = c("x", "y"), crs = "EPSG:4326")
```

```{r}
world_random = world[random_points, ]
nrow(world_random)
```

```{r}
random_joined = st_join(random_points, world["name_long"])
```

```{r}
#jm0 = tm_shape(world) + tm_borders(lwd = 0.2) + tm_format("World")

jm0 = tm_shape(world) + 
  tm_borders(lwd = 0.2) + 
  tm_layout(inner.margins=c(0, 0.05, 0.025, 0.01), 
            legend.position=tm_pos_in("left", "bottom"),
            component.position=c("right", "bottom"), 
            scale=.8, title.size = 1.3) + tm_crs("auto")

jm1 = jm0 +
  tm_shape(shp = random_points, bbox = bb) +
  tm_symbols(col = "black", shape = 4, lwd = 3)

jm2 = jm0 +
  tm_shape(world_random, bbox = bb) +
  tm_fill(fill = "name_long", fill.scale = tm_scale(values = "brewer.dark2")) +
  tm_layout(legend.show = FALSE)

jm3 = jm0 +
  tm_shape(shp = random_joined, bbox = bb) +
  tm_symbols(col = "name_long", shape = 4, lwd = 3, col.scale = tm_scale(values = "brewer.dark2")) +
  tm_layout(legend.show = FALSE)

jm4 = jm0 +
  tm_shape(shp = random_joined, bbox = bb) +
  tm_symbols(col = "name_long", shape = 4, lwd = 3, col.scale = tm_scale(values = "brewer.dark2")) +
  tm_layout(legend.only = TRUE)

tmap_arrange(jm1, jm2, jm3, jm4, nrow = 2, ncol = 2)
```

## Distance based joins

```{r}
plot(st_geometry(cycle_hire), col = "blue")
plot(st_geometry(cycle_hire_osm), add = TRUE, pch = 3, col = "red")
```

```{r}
any(st_intersects(cycle_hire, cycle_hire_osm, sparse = F))
```

> Find the points within 20m of each other

```{r}
sel = st_is_within_distance(cycle_hire, cycle_hire_osm, 
                            dist = units::set_units(20, "m"))
summary(lengths(sel) > 0)
```

```{r}
z = st_join(cycle_hire, cycle_hire_osm, 
            st_is_within_distance,
            dist = units::set_units(20, "m"))
nrow(z)
nrow(cycle_hire)
```

> Note that there are additional rows because there were some points with multiple matches, so:

```{r}
z = z |> 
  group_by(id) |> 
  summarize(capacity = mean(capacity))
nrow(z) == nrow(cycle_hire)
```

```{r}
plot(cycle_hire["name"])
plot(cycle_hire_osm["capacity"])
plot(z["capacity"])
```

## Spatial aggregation

> Aggregate values by a different geometry

```{r}
nz_agg = aggregate(x = nz_height, by = nz, FUN=mean)
```

> Alternatively

```{r}
nz_agg2 = st_join(x = nz, y = nz_height) |> 
  group_by(Name) |> 
  summarise(elevation = mean(elevation, na.rm=T))
```

tm_shape(nz_agg) + tm_fill("elevation", fill.scale = tm_scale(breaks = seq(27, 30, by = 0.5) \* 1e2)) + tm_borders() + tm_layout(scale = 1.8)

```{r}
tm_shape(nz_agg) +
  tm_fill("elevation", fill.scale=tm_scale(breaks=seq(27,30,by=0.5)*1e2)) +
  tm_borders() +
  tm_layout(scale = 1.8) +
  tm_crs("auto")
```

## Incongruent Areas

```         
rx = rbind(congruent, incongruent)
# tmap_mode("plot")
m1 = tm_shape(rx) +
  tm_fill("value", fill.scale = tm_scale(breaks = seq(3.5, 7, by = 0.5))) +
  tm_borders(lwd = 1, col = "black", lty = 1) +
  tm_facets(by = "level", drop.units = TRUE, ncol = 2) +
  tm_shape(aggregating_zones) +
  tm_borders(col_alpha = 0.2, lwd = 8, col = "red") +
  tm_layout(legend.show = FALSE, scale = 1)
m1
```

```{r}
rx = rbind(congruent, incongruent)
m1 = tm_shape(rx) +
  tm_fill("value", fill.scale = tm_scale(breaks = seq(3.5, 7, by = 0.5))) +
  tm_borders(lwd = 1, col = "black", lty = 1) +
  tm_facets(by = "level", drop.units = T, ncol = 2) +
  tm_shape(aggregating_zones) +
  tm_borders(col_alpha = 0.2, lwd = 8, col = "red") +
  tm_layout(legend.show = F, scale = 1)
m1
```

> Area-weighted spatial interpolation. `extensive = TRUE` uses `sum()`. Setting `extensive` to False to use `mean()` (intensive interpolation).

```{r}
iv = incongruent["value"]
agg_aw = st_interpolate_aw(iv, aggregating_zones, extensive = TRUE)
agg_aw$value
```
