---
title: "04SpatialDataOperations"
output: html_document
date: "2025-03-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Spatial Data Operations - Raster Data

```{r}
library(sf)
library(terra)
library(spData)
library(dplyr)
library(tmap)
```

```{r}
elev = rast(system.file("raster/elev.tif", package = "spData"))
grain = rast(system.file("raster/grain.tif", package = "spData"))
```

## Spatial subsetting

```{r}
id = cellFromXY(elev, xy = matrix(c(0.1, 0.1), ncol = 2))
elev[id]
```

> Alternatively

```{r}
terra::extract(elev, matrix(c(0.1,0.1), ncol = 2))
```

> Clipping with a raster object

```{r}
clip = rast(xmin = 0.9, xmax = 1.8, 
            ymin = -0.45, ymax = 0.45, 
            resolution = 0.3, vals = rep(1, 9))
elev[clip]
```

## NDVI example

The calculation of the normalized difference vegetation index (NDVI) is a well-known local (pixel-by-pixel) raster operation. It returns a raster with values between -1 and 1; positive values indicate the presence of living plants (mostly \> 0.2). NDVI is calculated from red and near-infrared (NIR) bands of remotely sensed imagery, typically from satellite systems such as Landsat or Sentinel. Vegetation absorbs light heavily in the visible light spectrum, and especially in the red channel, while reflecting NIR light. Here's the NDVI formula:

$$NDVI=\frac{NIR−Red}{NIR+Red}$$

Let's calculate NDVI for the multi-spectral satellite file of Zion National Park.

```{r}
multi_raster_file = system.file("raster/landsat.tif", package = "spDataLarge")
multi_rast = rast(multi_raster_file)
```

Our raster object has four satellite bands from the Landsat 8 satellite: blue, green, red, and NIR. Importantly, Landsat level-2 products are stored as integers to save disk space, and thus we need to convert them to floating-point numbers before doing any calculations. For that purpose, we need to apply a scaling factor (0.0000275) and add an offset (-0.2) to the original values.

```{r}
multi_rast = (multi_rast * 0.0000275) - 0.2
```

The proper values now should be in a range between 0 and 1. This is not the case here, probably due to the presence of clouds and other atmospheric effects, which are stored as negative values. We will replace these negative values with 0 as follows.

```{r}
multi_rast[multi_rast < 0] = 0
```

Implement the NDVI formula:

```{r}
ndvi_fun = function(nir, red){
  (nir - red) / (nir + red)
}
```

This function accepts two numerical arguments, `nir` and `red`, and returns a numerical vector with NDVI values. It can be used as the `fun` argument of [`lapp()`](https://rspatial.github.io/terra/reference/lapp.html). We just need to remember that our function expects two bands (not four from the original raster), and they need to be in the NIR, red order. That is why we subset the input raster with `multi_rast[[c(4, 3)]]` before doing any calculations.

```{r}
ndvi_rast = lapp(multi_rast[[c(4, 3)]], fun = ndvi_fun)
```
