---
title: "05.2-GeometricOperationsVectors"
output: html_document
date: "2025-03-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(sf)
library(terra)
library(dplyr)
library(spData)
library(spDataLarge)
library(tmap)
```

# Geometry Operations - Raster

## Intersections

> keep the matrix structure after clipping by setting `drop = FALSE`

```{r}
elev = rast(system.file("raster/elev.tif", package = "spData"))
clip = rast(xmin = 0.9, xmax = 1.8, ymin = -0.45, ymax = 0.45,
            resolution = 0.3, vals = rep(1, 9))
elev[clip, drop = FALSE]
```

## Extent and origin

```{r}
elev_2 = extend(elev, c(1,2))
```

```{r}
elev_poly = st_as_sf(as.polygons(elev, dissolve=F))
elev_2_poly = st_as_sf(as.polygons(elev_2, na.rm = F, dissolve = F))

tm1 = tm_shape(elev_poly, bbox = elev_2_poly) +
  tm_polygons(fill = "elev") +
  tm_layout(frame = F, legend.show = F)

tm2 = tm_shape(elev_2_poly) +
  tm_polygons(fill = "elev") +
  tm_layout(frame = F, legend.show = F)

tmap_arrange(tm1, tm2, nrow = 1)
```

```         
elev_3 = elev + elev_2
#> Error: [+] extents do not match
```

```{r}
elev_4 = extend(elev, elev_2)
origin(elev_4)
```

```{r}
origin(elev_4) = c(0.25, 0.25)
```

```{r}
elev_poly = st_as_sf(as.polygons(elev, dissolve = FALSE))
elev4_poly = st_as_sf(as.polygons(elev_4, dissolve = FALSE))

tm_shape(elev4_poly) +
  tm_grid() +
  tm_polygons(fill = "elev", lwd = 0.5) +
  tm_shape(elev_poly) +
  tm_polygons(fill = "elev") +
  tm_layout(frame = FALSE, legend.show = FALSE,
            inner.margins = c(0.1, 0.12, 0, 0))
```

## Aggregation and disaggregation

```{r}
dem = rast(system.file("raster/dem.tif", package = "spDataLarge"))
dem_agg = aggregate(dem, fact = 5, fun = mean)
```

```{r}
p_ar1 = tm_shape(dem) +
  tm_raster(col.scale = tm_scale_continuous()) +
  tm_title("A. Original") + 
  tm_layout(frame = FALSE, legend.show = FALSE)
p_ar2 = tm_shape(dem_agg) +
  tm_raster(col.scale = tm_scale_continuous()) +
  tm_title("B. Aggregated") +
  tm_layout(frame = FALSE, legend.show = FALSE)
tmap_arrange(p_ar1, p_ar2, ncol = 2)
```

```{r}
dem_disagg = disagg(dem_agg, fact = 5, method = "bilinear")
identical(dem, dem_disagg)
```

## Resampling

> When two rasters have different resolutions and origins

There are several methods for estimating values for a raster with different resolutions/origins, as shown in Figure [5.18](https://r.geocompx.org/geometry-operations#fig:resampl). The main resampling methods include:

-   Nearest neighbor: assigns the value of the nearest cell of the original raster to the cell of the target one. This is a fast simple technique that is usually suitable for resampling categorical rasters.

-   Bilinear interpolation: assigns a weighted average of the four nearest cells from the original raster to the cell of the target one (Figure [5.16](https://r.geocompx.org/geometry-operations#fig:bilinear)). This is the fastest method that is appropriate for continuous rasters.

-   Cubic interpolation: uses values of the 16 nearest cells of the original raster to determine the output cell value, applying third-order polynomial functions. Used for continuous rasters and results in a smoother surface compared to bilinear interpolation but is computationally more demanding.

-   Cubic spline interpolation: also uses values of the 16 nearest cells of the original raster to determine the output cell value, but applies cubic splines (piece-wise third-order polynomial functions). Used for continuous rasters.

-   Lanczos windowed sinc resampling: uses values of the 36 nearest cells of the original raster to determine the output cell value. Used for continuous rasters.^2^

To apply resampling, the **terra** package provides a [`resample()`](https://rspatial.github.io/terra/reference/resample.html) function. It accepts an input raster (`x`), a raster with target spatial properties (`y`), and a resampling method (`method`).

```{r}
target_rast = rast(xmin = 794650, xmax = 798250, 
                   ymin = 8931750, ymax = 8935350,
                   resolution = 300, crs = "EPSG:32717")
dem_resampl = resample(dem, y = target_rast, method = "bilinear")
```

```{r}
dem_resampl1 = resample(dem, target_rast, method = "near")
dem_resampl2 = resample(dem, target_rast, method = "bilinear")
dem_resampl3 = resample(dem, target_rast, method = "cubic")
dem_resampl4 = resample(dem, target_rast, method = "cubicspline")
dem_resampl5 = resample(dem, target_rast, method = "lanczos")

library(tmap)
tm1 = tm_shape(dem) +
  tm_raster(col.scale = tm_scale(breaks = seq(200, 1100, by = 150))) +
  tm_title("Original raster") +
  tm_layout(frame = FALSE, legend.show = FALSE)
tm2 = tm_shape(dem_resampl1) +
  tm_raster(col.scale = tm_scale(breaks = seq(200, 1100, by = 150))) +
  tm_title("near") +
  tm_layout(frame = FALSE, legend.show = FALSE)
tm3 = tm_shape(dem_resampl2) +
  tm_raster(col.scale = tm_scale(breaks = seq(200, 1100, by = 150))) +
  tm_title("bilinear") +
  tm_layout(frame = FALSE, legend.show = FALSE)
tm4 = tm_shape(dem_resampl3) +
  tm_raster(col.scale = tm_scale(breaks = seq(200, 1100, by = 150))) +
  tm_title("cubic") +
  tm_layout(frame = FALSE, legend.show = FALSE)
tm5 = tm_shape(dem_resampl4) +
  tm_raster(col.scale = tm_scale(breaks = seq(200, 1100, by = 150))) +
  tm_title("cubicspline") +
  tm_layout(frame = FALSE, legend.show = FALSE)
tm6 = tm_shape(dem_resampl5) +
  tm_raster(col.scale = tm_scale(breaks = seq(200, 1100, by = 150))) +
  tm_title("lanczos") +
  tm_layout(frame = FALSE, legend.show = FALSE)
tmap_arrange(tm1, tm2, tm3, tm4, tm5, tm6)
```
