---
title: "06-RasterVectorInteractions"
output: html_document
date: "2025-03-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(sf)
library(terra)
library(dplyr)
library(tmap)
library(rcartocolor)
library(spData)
library(ggplot2)
library(grid)
```

# Raster Vector Interactions

## Raster cropping

We will use two objects to illustrate raster cropping:

-   A `SpatRaster` object `srtm` representing elevation (meters above sea level) in southwestern Utah

-   A vector (`sf`) object `zion` representing Zion National Park

```{r}
srtm = rast(system.file("raster/srtm.tif", package="spDataLarge"))
zion = read_sf(system.file("vector/zion.gpkg", package="spDataLarge"))
zion = st_transform(zion, st_crs(srtm))
```

> `crop()` reduces extent, `mask()` replaces values outside with `NA`

```{r}
srtm_cropped = crop(srtm, zion)
srtm_masked = mask(srtm, zion)
```

> They are typically used together

```{r}
srtm_final = mask(srtm_cropped, zion)
```

To mask everything *inside* the bounds

```{r}
srtm_inv_masked = mask(srtm, zion, inverse=TRUE)
```

```{r}
terrain_colors = carto_pal(7, "Geyser")
pz1 = tm_shape(srtm) +
  tm_raster(col.scale = tm_scale_continuous(values = terrain_colors)) +
  tm_shape(zion) +
  tm_borders(lwd = 2) +
  tm_title("Original") +
  tm_layout(legend.show = F)
pz2 = tm_shape(srtm_cropped) +
  tm_raster(col.scale = tm_scale_continuous(values = terrain_colors)) +
  tm_shape(zion) +
  tm_borders(lwd = 2) +
  tm_title("Cropped") +
  tm_layout(legend.show = F)
pz3 = tm_shape(srtm_masked) +
  tm_raster(col.scale = tm_scale_continuous(values = terrain_colors)) +
  tm_shape(zion) +
  tm_borders(lwd = 2) +
  tm_title("Masked") +
  tm_layout(legend.show = F)
pz4 = tm_shape(srtm_inv_masked) +
  tm_raster(col.scale = tm_scale_continuous(values = terrain_colors)) +
  tm_shape(zion) +
  tm_borders(lwd = 2) +
  tm_title("Inverse mask") +
  tm_layout(legend.show = F)
tmap_arrange(pz1, pz2, pz3, pz4, ncol = 4, asp = NA)
```

## Raster extraction

Identifying and returning values associated with a target raster based on a geographic selector object.

The basic example is of extracting the value of a raster cell at specific **points**. For this purpose, we will use `zion_points`, which contain a sample of 30 locations within Zion National Park (Figure [6.2](https://r.geocompx.org/raster-vector#fig:pointextr)).

```{r}
data("zion_points", package = "spDataLarge")
elevation = terra::extract(srtm, zion_points)
zion_points = cbind(zion_points, elevation)
```

```{r}
tm1 = tm_shape(srtm) +
  tm_raster(col.scale = tm_scale_continuous(values = terrain_colors), 
            col.legend = tm_legend("Elevation (m above sea level")) +
  tm_shape(zion) +
  tm_borders(lwd = 2) +
  tm_shape(zion_points) +
  tm_symbols(fill = "black", size = 0.5) +
  tm_add_legend(type = "symbols", fill = "black", size = 0.7, 
                labels = "zion_points", shape = 21)
tm1
```

> While *line* selectors can be used, it is easier for processing to split the line into multiple points and extract the values of the points.

```{r}
zion_transect = cbind(c(-113.2, -112.9), c(37.45, 37.2)) |>
  st_linestring() |> 
  st_sfc(crs = crs(srtm)) |>
  st_sf(geometry = _)
```

```{r}
zion_transect$id = 1:nrow(zion_transect)
zion_transect = st_segmentize(zion_transect, dfMaxLength = 250)
zion_transect = st_cast(zion_transect, "POINT")
```

> Derive the distance between the first point and all subsequent points

```{r}
zion_transect = zion_transect |> 
  group_by(id) |> 
  mutate(dist = st_distance(geometry)[, 1])
```

> Finally, we can extract elevation values for each point in our transects and combine this information with our main object.

```{r}
zion_elev = terra::extract(srtm, zion_transect)
zion_transect = cbind(zion_transect, zion_elev)
```

```{r}
zion_transect_line = cbind(c(-113.2, -112.9), c(37.45, 37.2)) |>
  st_linestring() |> 
  st_sfc(crs = crs(srtm)) |>
  st_sf()
zion_transect_points = st_cast(zion_transect, "POINT")[c(1, nrow(zion_transect)), ]
zion_transect_points$name = c("start", "end")
```

```{r}
zion_transect_points
```

```{r}
rast_poly_line = tm_shape(srtm) +
  tm_raster(col.scale = tm_scale_continuous(values = terrain_colors),
            col.legend = tm_legend("Elevation (m")) + 
  tm_shape(zion) + 
  tm_borders(lwd = 2) +
  tm_shape(zion_transect_line) + 
  tm_lines(col = "black", lwd = 4) + 
  tm_shape(zion_transect_points) +
  tm_text("name", 
          text.scale = tm_scale(bg.color = "white", bg.alpha = 0.75, 
                                auto.placement = T)) +
  tm_layout(legend.frame = TRUE, legend.position = c("RIGHT", "TOP"),
            legend.bg.color = "white")

plot_transect = ggplot(zion_transect, aes(as.numeric(dist), srtm)) +
  geom_line() + 
  labs(x = "Distance (m)", y = "Elevation m asl") +
  theme_bw() +
  theme(plot.margin = unit(c(5.5, 15.5, 5.5, 5.5), "pt"))

grid.newpage()
pushViewport(viewport(
          layout = grid.layout(2, 2, heights = unit(c(0.25, 5), "null"))))
grid.text("A. Line extraction", 
          vp = viewport(layout.pos.row = 1, layout.pos.col = 1))
grid.text("B. Elevation along the line", 
          vp = viewport(layout.pos.row = 1, layout.pos.col = 2))
print(rast_poly_line, 
      vp = viewport(layout.pos.row = 2, layout.pos.col = 1))
print(plot_transect, 
      vp = viewport(layout.pos.row = 2, layout.pos.col = 2))

```

## Polygon extraction

```{r}
zion_srtm_values = terra::extract(x = srtm, y = zion)
```

```{r}
zion_srtm_values |> 
  group_by(ID) |> 
  summarise(across(srtm, list(min = min, 
                              mean = mean, 
                              max = max)))
```

```{r}
nlcd = rast(system.file("raster/nlcd.tif", 
                        package = "spDataLarge"))
zion2 = st_transform(zion, st_crs(nlcd))
zion_nlcd = terra::extract(nlcd, zion2)
zion_nlcd |> 
  group_by(ID, levels) |> 
  count()
```

```{r}
rast_poly_srtm = tm_shape(srtm) + 
  tm_raster(
    col.scale = tm_scale_continuous(values = terrain_colors),
    col.legend = tm_legend("Elevation (m)")) + 
  tm_shape(zion) +
  tm_polygons(lwd = 2, fill_alpha = 0.3) +
  tm_title("A. Continuous data extraction") +
  tm_layout(legend.frame = TRUE, 
            legend.position = c("LEFT", "BOTTOM"),
            legend.bg.color = "white")
rast_poly_nlcd = tm_shape(nlcd) +
  tm_raster(col.scale = tm_scale_categorical(levels.drop = TRUE),
            col.legend = tm_legend(title = "Land cover")) + 
  # tm_raster(drop.levels = TRUE, title = "Land cover", legend.show = TRUE) + 
  tm_shape(zion) +
  tm_polygons(lwd = 2, fill_alpha = 0.3)  +
  tm_title("B. Categorical data extraction") +
  tm_layout(legend.frame = TRUE, 
            legend.position = c("LEFT", "BOTTOM"),
            legend.bg.color = "white")
tmap_arrange(rast_poly_srtm, rast_poly_nlcd, ncol = 2)
```

## Rasterization of Vector Data

> Requires a template raster

To demonstrate rasterization in action, we will use a template raster that has the same extent and CRS as the input vector data `cycle_hire_osm_projected` (a dataset on cycle hire points in London is illustrated in Figure [6.5](https://r.geocompx.org/raster-vector#fig:vector-rasterization1)(A)) and spatial resolution of 1000 meters:

```{r}
cycle_hire_osm = spData::cycle_hire_osm
cycle_hire_osm_projected = st_transform(cycle_hire_osm, "EPSG:27700")
raster_template = rast(ext(cycle_hire_osm_projected), resolution = 1000,
                       crs = crs(cycle_hire_osm_projected))
```

### Points

> First approach
>
> Create a raster representing presence or absence of hire points

```{r}
ch_raster1 = rasterize(cycle_hire_osm_projected, raster_template)
```

> Second approach
>
> Use a function to summarize data in a raster cell, eg. number of cycle points in a cell using `length`

```{r}
ch_raster2 = rasterize(cycle_hire_osm_projected, raster_template, 
                       fun = "length")
```

> Third approach
>
> Each point has a `capacity`, which can be summed

```{r}
ch_raster3 = rasterize(cycle_hire_osm_projected, raster_template, 
                       field = "capacity", fun = sum, na.rm = T)
```

```{r}
r0p = tm_shape(cycle_hire_osm_projected) + 
  tm_symbols(fill = "capacity", size = 0.3, 
             fill.legend = tm_legend("Capacity: ")) + 
  tm_title("A. Points") +
  tm_layout(legend.position = c("RIGHT", "BOTTOM"), legend.frame = TRUE, 
  inner.margins = c(0.02, 0.02, 0.02, 0.35))

r1p = tm_shape(ch_raster1) + 
  tm_raster(col.scale = tm_scale_categorical(values = cols4all::c4a(n = 1, "hcl.blues3", reverse = TRUE)),
            col.legend = tm_legend("Values: ")) + 
  tm_title("B. Presence/absence") +
  tm_layout(legend.position = c("RIGHT", "BOTTOM"), legend.frame = TRUE, 
  inner.margins = c(0.02, 0.02, 0.02, 0.35))

r2p = tm_shape(ch_raster2) + 
  tm_raster(col.legend = tm_legend("Values: ")) + 
  tm_title("C. Count") +
  tm_layout(legend.position = c("RIGHT", "BOTTOM"), legend.frame = TRUE, 
  inner.margins = c(0.02, 0.02, 0.02, 0.35))

r3p = tm_shape(ch_raster3) + 
  tm_raster(col.legend = tm_legend("Values: ")) + 
  tm_title("D. Aggregated capacity") +
  tm_layout(legend.position = c("RIGHT", "BOTTOM"), legend.frame = TRUE, 
  inner.margins = c(0.02, 0.02, 0.02, 0.35))

tmap_arrange(r0p, r1p, r2p, r3p, ncol = 2)
```

### Lines

```{r}
california = dplyr::filter(us_states, NAME == "California")
california
```

```{r}
california_borders = st_cast(california, "MULTILINESTRING")
raster_template2 = rast(ext(california), resolution = 0.5,
                        crs = st_crs(california)$wkt)
```

```{r}
california_raster1 = rasterize(california_borders, raster_template2,
                               touches = TRUE)
```

## Spatial vectorization - raster to vector

### To points

```{r}
elev = rast(system.file("raster/elev.tif", package = "spData"))
elev_point = as.points(elev) |> 
  st_as_sf()
```

```{r}
p1 = tm_shape(elev) + 
  tm_raster(col.scale = tm_scale(n = 36)) +
  tm_title("A. Raster") +
  tm_layout(outer.margins = rep(0.01, 4), inner.margins = rep(0, 4), legend.show = FALSE)
p2 = tm_shape(elev_point) +
  tm_symbols(fill = "elev", fill.scale = tm_scale(n = 36), size = 2) +
  tm_title("B. Points") +
  tm_layout(outer.margins = rep(0.01, 4), inner.margins = rep(0.09, 4), legend.show = FALSE)
tmap_arrange(p1, p2, ncol = 2)
```

###  To lines

```{r}
dem = rast(system.file("raster/dem.tif", package = "spDataLarge"))
cl = as.contour(dem) |> 
  st_as_sf()
```

```{r}
plot(dem, axes = F)
plot(cl, add = T)
```

```{r}
hs = shade(slope = terrain(dem, "slope", unit = "radians"),
            aspect = terrain(dem, "aspect", unit = "radians"))
plot(hs, col = gray(0:100 / 100), legend = F)
plot(dem, col = terrain.colors(25), alpha = 0.5, legend = FALSE, add = TRUE)
contour(dem, col = "white", add = TRUE)
```

### To Polygons

```{r}
grain = rast(system.file("raster/grain.tif", package = "spData"))
grain_poly = as.polygons(grain) |> 
  st_as_sf()
```

```{r}
grain_poly = as.polygons(grain, dissolve = F) |> 
  st_as_sf()
grain_poly2 = as.polygons(grain) |> 
  st_as_sf()
cols = c("clay" = "brown", "sand" = "rosybrown", "silt" = "sandybrown")

p1p = tm_shape(grain) +
  tm_raster("grain", col.scale = tm_scale(values = cols)) +
  tm_title("A. Raster") +
  tm_layout(frame = FALSE, legend.show = FALSE)

p2p = tm_shape(grain_poly) +
  tm_polygons("grain", fill.scale = tm_scale(values = cols), lwd = 3) +
  tm_title("B. Polygons") +
  tm_layout(frame = FALSE, legend.show = FALSE)

p3p = tm_shape(grain_poly2) + 
  tm_polygons("grain", fill.scale = tm_scale(values = cols), lwd = 3)  +
  tm_title("C. Aggregated polygons") +
  tm_layout(frame = FALSE, legend.show = FALSE)

tmap_arrange(p1p, p2p, p3p, ncol = 3)
```
